---
title: "BIN_Analysis"
output: html_document
date: "2026-06-01"
---
# Word Recognition Task
## Bilinguals Timecourse (Updated 7/20/26)

There are 16 different trial types for bilinguals:

-   quiet (8): stroller, hand, car, foot, perro, nariz, leche, ojo

-   noise (8): hand, foot, stroller, car, perro, nariz, ojo, leche

Object pairs:

- ojo/perro
- car/foot
- stroller/hand
- nariz/leche

Notes from analysis 6/17/26: Only including bilinguals over 12 months old (N=24)

Notes from analysis 7/20/26: Added 1 more bilingual over 12 mos old (N = 25)
```{r, include=FALSE}
library(tidyverse)

#Reading in the timecourse output and plotting the bilingual timecourse in R.
btime <- read_csv(
  "bilingual_timecourse.csv",
  col_names = TRUE
)

btime <- btime %>%
  rename(condition = 1)

btime_long <- btime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Bilinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
btime_long <- btime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    ),
    word = str_remove(condition, "^(quiet|noise)_"),
    language = case_when(
      word %in% c("stroller", "hand", "car", "foot") ~ "English",
      word %in% c("perro", "nariz", "leche", "ojo") ~ "Spanish"
    )
  )
```

```{r, include=FALSE}
btime_plot <- btime_long %>%
  group_by(time, quiet_noise, language) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(btime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  facet_wrap(~ language) +
  labs(
    title = "Bilingual Timecourse by Language N=25",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```
## Accuracy Analysis: baseline- corrected (Added 7/20/26)

SES info added to analyis pipeline 7/29/26
```{r ses-prep, include=FALSE}
library(tidyverse)

# Bracket -> ordinal rank (1 = lowest). Midpoints are also fine, but with
# ">$75,000" open-ended, ranks are the more defensible choice.
income_levels <- c("<$15,000", "$15,000-$25,000", "$25,000-$35,000",
                   "$35,000-$45,000", "$45,000-$55,000", "$55,000-$65,000",
                   "$65,000-$75,000", ">$75,000")

ses <- read_csv("predictors.csv", show_col_types = FALSE) %>%
  select(participant, `parent ed`, income) %>%          # ignore the empty Unnamed cols
  rename(Participant = participant, parent_ed = `parent ed`) %>%
  mutate(
    income_rank = as.integer(factor(str_trim(income), levels = income_levels)),
    # z-score education across the full sample so coefficients are per-SD
    parent_ed_z = as.numeric(scale(parent_ed))
  )

# Sanity checks -- run once, look at them, then trust the join
ses %>% summarise(across(c(parent_ed, income_rank), ~sum(is.na(.))))
ses %>% filter(is.na(parent_ed)) %>% pull(Participant)
# expected missing: M43EW, B17YT, B23RF, B29NRP, B31ZR
```

```{r}
library(tidyverse)

# Reads one window pass. Recomputes proportion from frame counts (cols 5 = Frames to
# Target, 6 = Frames to Distractor) so NaN/DROPPED rows are handled by the count filter,
# not by parsing the accuracy string. total >= 15 frames == 0.5 s of looking at 30 fps.
read_window <- function(path, value) {
  raw <- read.csv(path, header = TRUE, stringsAsFactors = FALSE)
  tibble(
    Participant = as.character(raw[[1]]),
    Trial.Number = raw[[2]],
    Trial.Type   = str_trim(as.character(raw[[4]])),
    ft = suppressWarnings(as.numeric(raw[[5]])),   # Frames to Target
    fd = suppressWarnings(as.numeric(raw[[6]]))    # Frames to Distractor
  ) %>%
    filter(!is.na(Participant), str_trim(Participant) != "") %>%   # drop blank separator rows
    mutate(total = ft + fd) %>%
    filter(!is.na(total), total >= 15) %>%                         # 0.5 s looking; removes DROPPED
    transmute(
      Participant,
      trial = parse_number(as.character(Trial.Number)),            # "Trial 5" -> 5
      Trial.Type,
      !!value := ft / total
    )
}

# Joins a test pass (60:180) with its baseline pass (0:36) and subtracts, per trial.
build_corrected <- function(test_path, base_path, exclude_types) {
  test <- read_window(test_path, "test_acc")
  base <- read_window(base_path, "base_acc")
  inner_join(test, base, by = c("Participant", "trial", "Trial.Type")) %>%   # valid in BOTH windows
    mutate(
      corrected_acc = test_acc - base_acc,
      quiet_noise = case_when(
        str_starts(Trial.Type, "quiet") ~ "Quiet",
        str_starts(Trial.Type, "noise") ~ "Noise"
      )
    ) %>%
    filter(!Trial.Type %in% exclude_types)
}
```

## Bilinguals Accuracy (Updated 7/20/26)

-   This data reflects looking time data for a 4 second window from 2-6 seconds
- baseline-corrected proportion is calculated using baseline from 0-1.2 seconds

### Build corrected data and labels
```{r, include=FALSE}
b_exclude <- c("baseline_1","baseline_2","baseline_3","baseline_4",
               "quiet_eye","noise_eye","noise_nose","quiet_nose",
               "noise_milk","noise_dog","quiet_dog","quiet_milk")

bcorr <- build_corrected("bwindow.csv", "bbaseline.csv", b_exclude) %>%
  mutate(
    word = str_remove(Trial.Type, "^(quiet|noise)_"),
    language = case_when(
      word %in% c("stroller","hand","car","foot") ~ "English",
      word %in% c("perro","nariz","leche","ojo")  ~ "Spanish"
    )
  )
```

### check how many trials survive the baseline requirement (run once, then decide):

```{r}
# The baseline window is only 36 frames, so requiring 15 (=0.5 s) is a HIGH bar there.
# See how much you lose to "no valid baseline" before trusting the corrected numbers.
b_test_n <- read_window("bwindow.csv", "test_acc") %>%
  filter(!Trial.Type %in% b_exclude) %>% nrow()
b_corr_n <- nrow(bcorr)
message("Bilingual trials with valid TEST window: ", b_test_n,
        " | with valid test AND baseline: ", b_corr_n,
        " (", round(100 * b_corr_n / b_test_n), "% retained)")

```
### inclusion rules

```{r}
MIN_PER_COND <- 4   # >=4 valid corrected Quiet AND >=4 valid corrected Noise trials

keep_ids <- bcorr %>%
  count(Participant, quiet_noise) %>%
  pivot_wider(names_from = quiet_noise, values_from = n, values_fill = 0) %>%
  filter(Quiet >= MIN_PER_COND, Noise >= MIN_PER_COND) %>%
  pull(Participant)

bcorr_clean <- bcorr %>% filter(Participant %in% keep_ids)
message("Kept ", length(keep_ids), " of ", n_distinct(bcorr$Participant), " participants")
setdiff(unique(bcorr$Participant), keep_ids)   # who was dropped & why
```
```{r ses-join-wordrec}
bcorr_ses <- bcorr_clean %>% left_join(ses, by = "Participant")

# who in the looking data failed to match a predictors row? (should be empty)
bcorr_ses %>% filter(is.na(parent_ed)) %>% distinct(Participant)
```

### participant-level means

```{r}
p_level <- bcorr_clean %>%
  group_by(Participant, quiet_noise, language) %>%
  summarise(acc = mean(corrected_acc), n = n(), .groups = "drop")

p_level %>%
  group_by(quiet_noise, language) %>%
  summarise(mean_corr = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")
```
### tests against 0 (the null is now "no shift from baseline"):

```{r}
p_overall <- bcorr_clean %>%
  group_by(Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop")
t.test(p_overall$acc, mu = 0)          # is target looking above pre-naming baseline?

by_cond_wide <- bcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)
t.test(by_cond_wide$Quiet, by_cond_wide$Noise, paired = TRUE)   # Quiet vs Noise
```
### split by language

```{r}
# Bilingual: language x condition, participant-level means, one-sample t vs 0 per cell.
bilingual_lang_cond <- bcorr_clean %>%
  group_by(language, quiet_noise, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(language, quiet_noise) %>%
  summarise(
    N       = n(),
    Mean    = mean(acc),
    SE      = sd(acc) / sqrt(n()),
    CI_low  = t.test(acc, mu = 0)$conf.int[1],
    CI_high = t.test(acc, mu = 0)$conf.int[2],
    t       = t.test(acc, mu = 0)$statistic,
    df      = t.test(acc, mu = 0)$parameter,
    p       = t.test(acc, mu = 0)$p.value,
    .groups = "drop"
  ) %>%
  mutate(across(c(Mean, SE, CI_low, CI_high, t), ~round(., 3)),
         p = round(p, 4)) %>%
  arrange(language, quiet_noise)

bilingual_lang_cond
```

### mixed effects model

```{r bilingual-mixed-ses}
library(lme4); library(lmerTest)

# Model WITHOUT SES, refit on the SES-complete subset so the comparison is
# apples-to-apples (same rows), then the model WITH SES.
b_ses_complete <- bcorr_ses %>% filter(!is.na(parent_ed_z))

m0 <- lmer(corrected_acc ~ quiet_noise * language + (1 | Participant),
           data = b_ses_complete)
m1 <- lmer(corrected_acc ~ quiet_noise * language + parent_ed_z + (1 | Participant),
           data = b_ses_complete)

summary(m1)
anova(m0, m1)   # does SES improve fit? (refits with ML for the LRT)
message("N participants in SES models: ", n_distinct(b_ses_complete$Participant))
```

```{r}
library(lme4); library(lmerTest)
m <- lmer(corrected_acc ~ quiet_noise * language + (1 | Participant), data = bcorr_clean)
summary(m)
```

### plot by object

```{r}
b_by_object <- bcorr_clean %>%
  group_by(language, word, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%   # participant means per object
  group_by(language, word) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

ggplot(b_by_object, aes(x = mean, y = reorder(word, mean), color = language)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_pointrange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.8) +
  labs(
    title = "Bilingual: Baseline-Corrected Accuracy by Word",
    x = "Corrected accuracy (target looking above baseline)",
    y = NULL, color = "Language"
  ) +
  theme_classic()
```
### good words :)

```{r}
good_words <- c("leche", "car", "perro", "hand")

b_good <- bcorr_clean %>% filter(word %in% good_words)

# participant-level means, word x condition (descriptive)
b_good_cell <- b_good %>%
  group_by(word, quiet_noise, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(word, quiet_noise) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

b_good_cell
```
```{r}
ggplot(b_good_cell, aes(x = quiet_noise, y = mean, fill = quiet_noise)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  geom_hline(yintercept = 0, color = "gray40") +
  facet_wrap(~ word) +
  labs(
    title = "Bilingual: Quiet vs. Noise for Recognized Words",
    subtitle = "Exploratory — words selected post-hoc for being above 0",
    x = NULL, y = "Corrected accuracy", fill = "Condition"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```
```{r}
b_good_wide <- b_good %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)

t.test(b_good_wide$Quiet, b_good_wide$Noise, paired = TRUE)   # exploratory
```

### Ignore for now
Plotting accuracy by participant and condition
```{r}
library(dplyr)
library(ggplot2)

# Calculate mean accuracy for each participant in each condition
acc_b <- imputed_b %>%
  group_by(Participant, quiet_noise) %>%
  summarize(
    Accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(acc_b,
       aes(x = quiet_noise,
           y = Accuracy,
           group = Participant)) +
  geom_line(alpha = .4) +
  geom_point(size = 2) +
  theme_classic() +
  labs(
    x = "Condition",
    y = "Mean Accuracy"
  )
```

Calculating a noise difference score (quiet - noise)
```{r}
diff_b <- acc_b %>%
  tidyr::pivot_wider(
    names_from = quiet_noise,
    values_from = Accuracy
  ) %>%
  mutate(
    NoiseDrop = Quiet - Noise
  ) %>%
  arrange(desc(NoiseDrop))

head(diff_b)
```

Plotting diff scores so we can see:
- participants who perform much worse in noise (large positive values),
- participants with little effect of noise (near 0),
- participants who oddly perform better in noise (negative values).

```{r}
ggplot(diff_b,
       aes(x = reorder(Participant, NoiseDrop),
           y = NoiseDrop)) +
  geom_point() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Participant",
    y = "Quiet - Noise Accuracy"
  )
```

## Monolinguals Timecourse (Updated 7/20/26)

There are 16 different trial types for monlinguals:

-   quiet (8): stroller, hand, car, foot, eye, nose, dog, milk

-   noise (8): hand, foot, stroller, car, eye, nose, milk, dog

Object pairs:

- eye/dog
- car/foot
- stroller/hand
- milk/nose

Notes from analysis 6/17/26: Only including monolinguals over 12 months old (N=28)

Notes from analysis 7/20/26: Added 2 more monolinguals over 12 mos (N = 30)
```{r, include=FALSE}
mtime <- read_csv(
  "monolingual_timecourse.csv",
  col_names = TRUE
)

mtime <- mtime %>%
  rename(condition = 1)

mtime_long <- mtime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Monolinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
mtime_long <- mtime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    )
  )
```

```{r, include=FALSE}
mtime_plot <- mtime_long %>%
  group_by(time, quiet_noise) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(mtime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
   title = "Monolingual Timecourse N=30",
   x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

## Monolinguals Accuracy (Updated 7/20/26)

### build corrected data
```{r}
# Monolinguals hear all 8 objects in English; drop only the baseline trials.
m_exclude <- c("baseline_1", "baseline_2", "baseline_3", "baseline_4")

mcorr <- build_corrected("mwindow.csv", "mbaseline.csv", m_exclude)
```

### baseline survival

```{r}
m_test_n <- read_window("mwindow.csv", "test_acc") %>%
  filter(!Trial.Type %in% m_exclude) %>% nrow()
m_corr_n <- nrow(mcorr)
message("Monolingual trials with valid TEST window: ", m_test_n,
        " | with valid test AND baseline: ", m_corr_n,
        " (", round(100 * m_corr_n / m_test_n), "% retained)")
```
### inclusion rules

```{r}
m_keep_ids <- mcorr %>%
  count(Participant, quiet_noise) %>%
  pivot_wider(names_from = quiet_noise, values_from = n, values_fill = 0) %>%
  filter(Quiet >= MIN_PER_COND, Noise >= MIN_PER_COND) %>%
  pull(Participant)

mcorr_clean <- mcorr %>% filter(Participant %in% m_keep_ids)
message("Kept ", length(m_keep_ids), " of ", n_distinct(mcorr$Participant), " participants")
setdiff(unique(mcorr$Participant), m_keep_ids)   # who was dropped & why
```
```{r}
# adding in ses info
mcorr_ses <- mcorr_clean %>% left_join(ses, by = "Participant")

# who in the looking data failed to match a predictors row? (should be empty)
mcorr_ses %>% filter(is.na(parent_ed)) %>% distinct(Participant)
```

### participant level means

```{r}
m_p_level <- mcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), n = n(), .groups = "drop")

m_p_level %>%
  group_by(quiet_noise) %>%
  summarise(mean_corr = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")
```
### tests against zero

```{r}
m_overall <- mcorr_clean %>%
  group_by(Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop")
t.test(m_overall$acc, mu = 0)          # is target looking above pre-naming baseline?

m_by_cond_wide <- mcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)
t.test(m_by_cond_wide$Quiet, m_by_cond_wide$Noise, paired = TRUE)   # Quiet vs Noise
```
### plot by object

```{r}
m_by_object <- mcorr_clean %>%
  mutate(word = str_remove(Trial.Type, "^(quiet|noise)_")) %>%   # mcorr_clean has no 'word' yet
  group_by(word, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(word) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

ggplot(m_by_object, aes(x = mean, y = reorder(word, mean))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_pointrange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.8) +
  labs(
    title = "Monolingual: Baseline-Corrected Accuracy by Word",
    x = "Corrected accuracy (target looking above baseline)",
    y = NULL
  ) +
  theme_classic()
```
### good words :)

```{r}
m_good_words <- c("milk", "hand", "car", "dog")

m_good <- mcorr_clean %>%
  mutate(word = str_remove(Trial.Type, "^(quiet|noise)_")) %>%   # mcorr_clean has no 'word' column
  filter(word %in% m_good_words)

# participant-level means, word x condition (descriptive)
m_good_cell <- m_good %>%
  group_by(word, quiet_noise, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(word, quiet_noise) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

m_good_cell
```
```{r}
ggplot(m_good_cell, aes(x = quiet_noise, y = mean, fill = quiet_noise)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  geom_hline(yintercept = 0, color = "gray40") +
  facet_wrap(~ word) +
  labs(
    title = "Monolingual: Quiet vs. Noise for Recognized Words",
    subtitle = "Exploratory — words selected post-hoc for being above 0",
    x = NULL, y = "Corrected accuracy", fill = "Condition"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```
```{r}
m_good_wide <- m_good %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)

t.test(m_good_wide$Quiet, m_good_wide$Noise, paired = TRUE)   # exploratory
```

### mixed model

```{r mono-mixed-ses}
m_ses_complete <- mcorr_ses %>% filter(!is.na(parent_ed_z))

mm0 <- lmer(corrected_acc ~ quiet_noise + (1 | Participant),
            data = m_ses_complete)
mm1 <- lmer(corrected_acc ~ quiet_noise + parent_ed_z + (1 | Participant),
            data = m_ses_complete)

summary(mm1)
anova(mm0, mm1)
message("N participants in SES models: ", n_distinct(m_ses_complete$Participant))
```

```{r}
library(lme4); library(lmerTest)
mm <- lmer(corrected_acc ~ quiet_noise + (1 | Participant), data = mcorr_clean)
summary(mm)
```
## Combined summary table

```{r}
library(broom)

# Helper: one-sample t vs 0 on participant-level corrected means, returned as a tidy row
summ_vs0 <- function(df, group_label) {
  p <- df %>%
    group_by(Participant) %>%
    summarise(acc = mean(corrected_acc), .groups = "drop")
  tt <- t.test(p$acc, mu = 0)
  tibble(
    Group   = group_label,
    N       = nrow(p),
    Mean    = unname(tt$estimate),
    CI_low  = tt$conf.int[1],
    CI_high = tt$conf.int[2],
    t       = unname(tt$statistic),
    df      = unname(tt$parameter),
    p       = tt$p.value
  )
}

summary_table <- bind_rows(
  summ_vs0(mcorr_clean, "Monolingual"),
  summ_vs0(bcorr_clean, "Bilingual")
) %>%
  mutate(across(c(Mean, CI_low, CI_high, t), ~round(., 3)),
         p = round(p, 4))

summary_table
```

## Combined model with good objects
```{r}
library(lme4); library(lmerTest); library(tidyverse)

# --- Canonical object labels for the 4 shared objects ---
# Bilinguals heard leche/perro in Spanish; monolinguals heard milk/dog in English —
# same physical objects, so collapse to one label per object.
to_object <- function(word) {
  recode(word,
         "leche" = "milk", "perro" = "dog",   # Spanish -> shared object label
         "milk"  = "milk", "dog"   = "dog",
         "hand"  = "hand", "car"   = "car",
         .default = NA_character_)             # everything else -> NA, dropped below
}

shared_objects <- c("milk", "dog", "hand", "car")

# --- Bilingual side: already has `word`; add object + group ---
b_matched <- bcorr_clean %>%
  mutate(object = to_object(word),
         group  = "Bilingual") %>%
  filter(object %in% shared_objects) %>%
  select(Participant, group, object, quiet_noise, corrected_acc)

# --- Monolingual side: derive word first, then object + group ---
m_matched <- mcorr_clean %>%
  mutate(word   = str_remove(Trial.Type, "^(quiet|noise)_"),
         object = to_object(word),
         group  = "Monolingual") %>%
  filter(object %in% shared_objects) %>%
  select(Participant, group, object, quiet_noise, corrected_acc)

# --- Combine (B/M prefixes keep Participant IDs unique across groups) ---
matched <- bind_rows(b_matched, m_matched) %>%
  mutate(
    group       = factor(group, levels = c("Monolingual", "Bilingual")),  # ref = Monolingual
    quiet_noise = factor(quiet_noise, levels = c("Quiet", "Noise")),       # ref = Quiet
    object      = factor(object),
    Participant = factor(Participant)
  )

# --- Sanity check the combined dataset BEFORE modeling ---
cat("Rows:", nrow(matched),
    "| Participants:", nlevels(matched$Participant),
    "| Objects:", nlevels(matched$object), "\n")
matched %>% count(group, object)                       # trials per object per group
matched %>% count(group, quiet_noise)                  # trials per condition per group
matched %>% distinct(Participant, group) %>% count(group)  # N per group
```
```{r}
m_matched_mod <- lmer(
  corrected_acc ~ group * quiet_noise + (1 | Participant) + (1 | object),
  data = matched
)
summary(m_matched_mod)
```

```{r}
library(emmeans)

emm <- emmeans(m_matched_mod, ~ group * quiet_noise)
emm                                   # the four estimated cell means (vs 0 = no recognition)

pairs(emm, simple = "group")          # Bilingual vs Monolingual, within each condition
pairs(emm, simple = "quiet_noise")    # Quiet vs Noise, within each group
```

## Combined Model with SES

```{r combined-mixed-ses}
matched_ses <- matched %>%
  left_join(ses, by = "Participant") %>%
  filter(!is.na(parent_ed_z))

# collinearity check: how separated are the groups on SES in this subset?
matched_ses %>% distinct(Participant, group, parent_ed) %>%
  group_by(group) %>%
  summarise(n = n(), mean_ed = mean(parent_ed), sd_ed = sd(parent_ed),
            min = min(parent_ed), max = max(parent_ed))

c0 <- lmer(corrected_acc ~ group * quiet_noise + (1 | Participant) + (1 | object),
           data = matched_ses)
c1 <- lmer(corrected_acc ~ group * quiet_noise + parent_ed_z +
             (1 | Participant) + (1 | object),
           data = matched_ses)

summary(c1)
anova(c0, c1)

# Check the variance inflation on the fixed effects; VIF >> 5 for group and
# parent_ed_z confirms they can't be separated in this sample.
car::vif(c1)
```

```{r ses-within-bilinguals}
b_noise_ses <- lmer(corrected_acc ~ quiet_noise * parent_ed_z + language +
                      (1 | Participant),
                    data = b_ses_complete)
summary(b_noise_ses)
# quiet_noise:parent_ed_z is the term of interest: does the Quiet-Noise gap
# shrink or grow with SES?
```

### Ignore for now
Plotting accuracy by participant and condition
```{r}
# Calculate mean accuracy for each participant in each condition
acc_m <- imputed_m %>%
  group_by(Participant, quiet_noise) %>%
  summarize(
    Accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(acc_m,
       aes(x = quiet_noise,
           y = Accuracy,
           group = Participant)) +
  geom_line(alpha = .4) +
  geom_point(size = 2) +
  theme_classic() +
  labs(
    x = "Condition",
    y = "Mean Accuracy"
  )

```

Calculating a noise difference score (quiet - noise)
```{r}
diff_m <- acc_m %>%
  tidyr::pivot_wider(
    names_from = quiet_noise,
    values_from = Accuracy
  ) %>%
  mutate(
    NoiseDrop = Quiet - Noise
  ) %>%
  arrange(desc(NoiseDrop))

head(diff_m)
```

Plotting diff scores so we can see:
- participants who perform much worse in noise (large positive values),
- participants with little effect of noise (near 0),
- participants who oddly perform better in noise (negative values).

```{r}
ggplot(diff_m,
       aes(x = reorder(Participant, NoiseDrop),
           y = NoiseDrop)) +
  geom_point() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Participant",
    y = "Quiet - Noise Accuracy"
  )
```

# Habituation Task
## Data Cleaning (updated 6/16/26)
- 79 participants (including V1 and V2 data points) with habituation data
this only includes kids who did habituate
- need to filter through kids and pick one data point (either V1 or V2) for every participant --> keep V2 point if there is a V2, get rid of their V1

```{r}
library(dplyr)
library(stringr)

# keep only 1 data point per participant, prioritizing a V2 if they have one
hab <- read.csv("hab.csv")

hab_filtered <- hab %>%
  mutate(
    base_id = str_remove(Participant, "_V[12]$"),          # strip visit suffix to get participant ID
    visit   = if_else(str_detect(Participant, "_V2$"), 2, 1)
  ) %>%
  group_by(base_id) %>%
  slice_max(visit, n = 1, with_ties = FALSE) %>%           # keep V2 if present, else V1
  ungroup()

# sanity checks
nrow(hab_filtered)                      # should equal number of unique participants
any(duplicated(hab_filtered$base_id))   # should be FALSE
table(hab_filtered$visit)               # how many V2s vs V1s were kept
```

```{r}
age  <- read.csv("age.csv")
lena <- read.csv("lena.csv")
mcdi <- read.csv("mcdi.csv")

hab_filtered <- hab_filtered %>%
  left_join(age,  by = "Participant") %>%
  left_join(lena, by = "Participant") %>%
  left_join(mcdi, by = "Participant") %>%
  left_join(ses, by = "Participant") %>%
  mutate(
    language = if_else(str_starts(base_id, "M"), "monolingual", "bilingual")
  )

hab_filtered %>% filter(is.na(parent_ed)) %>% pull(Participant)  # expect the 5 known-missing

```

### Missing Data
```{r}
#which kids are missing lena data
hab_filtered %>%
  filter(is.na(noise_sum)) %>%
  pull(Participant)

#19 kids 
#[1] "M05ES"     "M08SA"     "M43EW"     "M47JJ"     "M48DZ"    
#[6] "M08SA_V2"  "M24JB_V2"  "M37JD_V2"  "B04DR"     "B06JR"    
#[11] "B10KG"     "B17YT"     "B22JR"     "B24EBR"    "B23RF"    
#[16] "B29NRP"    "B31ZR"     "B32NF"     "B12KRM_V2"
```

```{r}
#which kids are missing mcdi data
hab %>%
  filter(is.na(mcdi)) %>%
  pull(Participant)

#32 kids 
#[1] "M11BLP"    "M22AS"     "M41GG"     "M43EW"     "M45DA"    
#[6] "M47JJ"     "M48DZ"     "M46CC"     "M12RO_V2"  "M13CY_V2" 
#[11] "M37JD_V2"  "B05ET"     "B10KG"     "B16LR"     "B17YT"    
#[16] "B18LS"     "B21HM"     "B22JR"     "B24EBR"    "B23RF"    
#[21] "B27KDS"    "B28APF"    "B30MR"     "B29NRP"    "B31ZR"    
#[26] "B34ES"     "B26GB"     "B32NF"     "B33KLH"    "B05ET_V2" 
#[31] "B11ARF_V2" "B12KRM_V2"
```

```{r}
# seeing how many kids have a V2 data point for habit data 
hab$visit <- ifelse(grepl("_V2$", hab$Participant), "V2", "V1")

table(hab$language, hab$visit)

#            V1 V2
#bilingual   28  3
#monolingual 32 16
```

### Age groups
```{r}
# creating age group bins
hab_filtered <- hab_filtered %>%
  mutate(
    age_group = case_when(
      age >= 8  & age < 13 ~ "8–12.9",
      age >= 13 & age < 19 ~ "13–18.9",
      age >= 19 & age < 27 ~ "19–26.9",
      TRUE ~ NA_character_
    ),
    age_group = factor(
      age_group,
      levels = c("8–12.9", "13–18.9", "19–26.9")
    )
  )

library(ggplot2)
ggplot(hab_filtered, aes(x = age)) +
  geom_histogram(binwidth = 1) +
  geom_rug() +
  theme_classic()
```

## Stats
### Linear regressions

```{r}
bi   <- hab_filtered %>% filter(language == "bilingual")
mono <- hab_filtered %>% filter(language == "monolingual")
```
```{r}
library(broom)
library(purrr)
library(dplyr)
library(tidyr)

specs <- crossing(
  outcome = c("nov_score", "hab_trials", "nov1"),
  nesting(
    predictor = c("age * language", "tv", "noise_sum"),
    sample    = c("full", "bilingual", "monolingual")
  )
)

datasets <- list(full = hab_filtered, bilingual = bi, monolingual = mono)

results <- specs %>%
  mutate(fit = pmap(list(outcome, predictor, sample),
                    \(o, p, s) lm(reformulate(p, response = o),
                                  data = datasets[[s]]))) %>%
  mutate(tidied = map(fit, tidy)) %>%
  unnest(tidied) %>%
  select(-fit)

results
```

```{r}
# just the predictor effects, sorted by p-value
results %>%
  filter(term != "(Intercept)") %>%
  arrange(p.value)
```
### linear regression with SES

```{r hab-specs-ses}
specs <- crossing(
  outcome = c("nov_score", "hab_trials", "nov1"),
  nesting(
    predictor = c("age * language",
                  "tv",
                  "noise_sum",
                  "parent_ed_z",                     # (a) SES alone
                  "age * language + parent_ed_z",    # (b) SES-adjusted versions
                  "tv + parent_ed_z",
                  "noise_sum + parent_ed_z"),
    sample    = rep(c("full", "bilingual", "monolingual"), length.out = 7)
  )
)
# NOTE: the nesting() above pairs each predictor with ONE sample the way your
# original did. If you instead want every predictor crossed with every sample:
specs <- crossing(
  outcome   = c("nov_score", "hab_trials", "nov1"),
  predictor = c("age * language", "tv", "noise_sum",
                "parent_ed_z",
                "age * language + parent_ed_z",
                "tv + parent_ed_z",
                "noise_sum + parent_ed_z"),
  sample    = c("full", "bilingual", "monolingual")
) %>%
  # drop nonsensical combos: language terms in single-group samples
  filter(!(str_detect(predictor, "language") & sample != "full"))

datasets <- list(full = hab_filtered, bilingual = bi, monolingual = mono)
# (rebuild bi / mono AFTER the SES join so they carry parent_ed_z)
bi   <- hab_filtered %>% filter(language == "bilingual")
mono <- hab_filtered %>% filter(language == "monolingual")
datasets <- list(full = hab_filtered, bilingual = bi, monolingual = mono)

results <- specs %>%
  mutate(fit = pmap(list(outcome, predictor, sample),
                    \(o, p, s) lm(reformulate(p, response = o),
                                  data = datasets[[s]]))) %>%
  mutate(n_used = map_int(fit, ~nobs(.x)),          # track N per model (SES missingness!)
         tidied = map(fit, tidy)) %>%
  unnest(tidied) %>%
  select(-fit)

results %>% filter(term != "(Intercept)") %>% arrange(p.value)
```

### confirmatory model based on exploratory grid

```{r}
library(broom)

# Confirmatory model: pre-specified after the exploratory grid.
# Full sample; SES entered as a covariate alongside the predictor of interest (tv).
confirm_df <- hab_filtered %>%
  filter(!is.na(nov_score), !is.na(tv), !is.na(age),
         !is.na(language), !is.na(parent_ed_z))

confirm_mod <- lm(nov_score ~ tv + age + language + parent_ed_z,
                  data = confirm_df)

summary(confirm_mod)
tidy(confirm_mod, conf.int = TRUE)      # coefficients with 95% CIs
glance(confirm_mod)                     # model R², F, df
nobs(confirm_mod)                       # report this N explicitly
```

## Plotting Novelty Response Score
```{r}
library(ggplot2)

ggplot(hab_filtered, aes(x = language, y = nov_score)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  labs(
    x = "Language Group",
    y = "Novelty Response Score",
    title = "Novelty Response Scores by Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab_filtered, aes(x = age, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Age (months)",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab_filtered, aes(x = age_group, y = nov_score, color = language)) +
  geom_boxplot(position = position_dodge(width = 0.8),
               outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.15,
    dodge.width = 0.8
  ),
  alpha = 0.7) +
  labs(
    x = "Age Group (months)",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

## Hab + LENA plots
```{r}
ggplot(hab_filtered, aes(x = noise_sum, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "LENA Total Noise",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab_filtered, aes(x = tv, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "LENA TV Noise",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```
