---
title: "BIN_Analysis"
output: html_document
date: "2026-06-01"
---

# Word Rec Task
## Bilinguals Timecourse

There are 16 different trial types for bilinguals:

-   quiet (8): stroller, hand, car, foot, perro, nariz, leche, ojo

-   noise (8): hand, foot, stroller, car, perro, nariz, ojo, leche

Object pairs:

-   ojo/perro
-   car/foot
-   stroller/hand
-   nariz/leche

```{r, include=FALSE}
library(tidyverse)

#Reading in the timecourse output and plotting the bilingual timecourse in R.
btime <- read_csv(
  "bilingual_timecourse.csv",
  col_names = TRUE
)

btime <- btime %>%
  rename(condition = 1)

btime_long <- btime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Bilinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
btime_long <- btime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    )
  )
```

```{r, include=FALSE}
btime_plot <- btime_long %>%
  group_by(time, quiet_noise) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(btime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
   title = "Bilingual Timecourse N=14",
   x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

Timecourse plot across all 16 conditions.

```{r, warning=FALSE}
ggplot(btime_long, aes(x = time, y = looking_time, color = condition)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
    title = "Bilingual Timecourse N=14",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition"
  ) +
  theme_classic()
```

I want to just plot all the quiet conditions in one plot because it's really messy to look at all 16 trial types.

```{r, warning=FALSE}
btime_quiet <- btime_long %>%
  filter(str_starts(condition, "quiet_"))

ggplot(btime_quiet, aes(x = time, y = looking_time, color = condition)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
    title = "Bilingual Quiet Conditions Timecourse",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition"
  ) +
  theme_classic()

```

Now just the noise conditions:

```{r, warning=FALSE}
btime_noise <- btime_long %>%
  filter(str_starts(condition, "noise_"))

ggplot(btime_noise, aes(x = time, y = looking_time, color = condition)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
    title = "Bilingual Noise Conditions Timecourse",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition"
  ) +
  theme_classic()
```

Ok let's look at this object by object:

```{r, warning=FALSE}
btime_object <- btime_long %>%
  mutate(object = str_remove(condition, "^(quiet_|noise_)"))

ggplot(btime_object, aes(x = time, y = looking_time, color = quiet_noise)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  facet_wrap(~ object, ncol = 4) +
  labs(
    title = "Bilingual Timecourse by Object (Quiet vs Noise)",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition Type"
  ) +
  theme_classic()
```

## Bilinguals Accuracy

-   This data reflects looking time data for a 2 second window (from 2-4 seconds)

-   After looking at the individual object plots (see above), I decided to remove the objects foot, hand, and leche because of low accuracy in the quiet (and noise) conditions.

```{r, include=FALSE}
bwindow <- read.csv("bwindow.csv", header=TRUE,na.strings=c(""))

#remove rows based on NA in Participant column (empty rows)
bwindow <- bwindow %>% drop_na(Participant)

#remove trials that we don't want based on the name
bwindow <- bwindow[!bwindow$Trial.Type %in% c("baseline_1", "baseline_2", "baseline_3", "baseline_4", "quiet_eye", "noise_eye", "quiet_nose", "noise_nose", "quiet_milk", "noise_milk", "quiet_dog", "noise_dog"),]

bwindow <- bwindow %>%
  mutate(
    quiet_noise = case_when(
      str_starts(Trial.Type, "quiet") ~ "Quiet",
      str_starts(Trial.Type, "noise") ~ "Noise"
    )
  )

rownames(bwindow) <- NULL
```

This table shows accuracy in the quiet and noise conditions (includes 10 different trial types - 5 objects named once in each condition).

```{r, warning=FALSE}
bwindow %>%
  group_by(quiet_noise) %>%
  summarise(
    mean_accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )
```

## Monolinguals Timecourse

There are 16 different trial types for monlinguals:

-   quiet (8): stroller, hand, car, foot, eye, nose, dog, milk

-   noise (8): hand, foot, stroller, car, eye, nose, milk, dog

Object pairs:

-   eye/dog
-   car/foot
-   stroller/hand
-   milk/nose

```{r, include=FALSE}
mtime <- read_csv(
  "monolingual_timecourse.csv",
  col_names = TRUE
)

mtime <- mtime %>%
  rename(condition = 1)

mtime_long <- mtime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Monolinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
mtime_long <- mtime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    )
  )
```

```{r, include=FALSE}
mtime_plot <- mtime_long %>%
  group_by(time, quiet_noise) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(mtime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
   title = "Monolingual Timecourse N=23",
   x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

Timecourse plot across all 16 conditions.

```{r, warning=FALSE}
ggplot(mtime_long, aes(x = time, y = looking_time, color = condition)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
    title = "Monolingual Timecourse N=23",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition"
  ) +
  theme_classic()
```

I want to just plot all the quiet conditions in one plot because it's really messy to look at all 16 trial types.

```{r, warning=FALSE}
mtime_quiet <- mtime_long %>%
  filter(str_starts(condition, "quiet_"))

ggplot(mtime_quiet, aes(x = time, y = looking_time, color = condition)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
    title = "Monolingual Quiet Conditions Timecourse",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition"
  ) +
  theme_classic()
```

Now just the noise conditions:

```{r, warning=FALSE}
mtime_noise <- mtime_long %>%
  filter(str_starts(condition, "noise_"))

ggplot(mtime_noise, aes(x = time, y = looking_time, color = condition)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
    title = "Monolingual Noise Conditions Timecourse",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition"
  ) +
  theme_classic()
```

Ok let's look at this object by object:

```{r, warning=FALSE}
mtime_object <- mtime_long %>%
  mutate(object = str_remove(condition, "^(quiet_|noise_)"))

ggplot(mtime_object, aes(x = time, y = looking_time, color = quiet_noise)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  facet_wrap(~ object, ncol = 4) +
  labs(
    title = "Monolingual Timecourse by Object (Quiet vs Noise)",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Condition Type"
  ) +
  theme_classic()
```

## Monolinguals Accuracy

-   I ran the same window for monolinguals (2-4 seconds)

-   I am going to remove the same 3 words that I did for the bilinguals (hand, foot, milk)

```{r, include=FALSE}
mwindow <- read.csv("mwindow.csv", header=TRUE,na.strings=c(""))

#remove rows based on NA in Participant column (empty rows)
mwindow <- mwindow %>% drop_na(Participant)

#remove trials that we don't want based on the name
mwindow <- mwindow[!mwindow$Trial.Type %in% c("baseline_1", "baseline_2", "baseline_3", "baseline_4", "quiet_ojo","noise_perro", "noise_nariz", "quiet_perro", "quiet_nariz", "noise_ojo", "quiet_leche", "noise_leche"),]

mwindow <- mwindow %>%
  mutate(
    quiet_noise = case_when(
      str_starts(Trial.Type, "quiet") ~ "Quiet",
      str_starts(Trial.Type, "noise") ~ "Noise"
    )
  )

rownames(mwindow) <- NULL
```

```{r, warning=FALSE}
mwindow %>%
  group_by(quiet_noise) %>%
  summarise(
    mean_accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )
```