---
title: "BIN_Analysis"
output: html_document
date: "2026-06-01"
---
# Word Recognition Task
## Bilinguals Timecourse (Updated 7/20/26)

There are 16 different trial types for bilinguals:

-   quiet (8): stroller, hand, car, foot, perro, nariz, leche, ojo

-   noise (8): hand, foot, stroller, car, perro, nariz, ojo, leche

Object pairs:

- ojo/perro
- car/foot
- stroller/hand
- nariz/leche

Notes from analysis 6/17/26: Only including bilinguals over 12 months old (N=24)

Notes from analysis 7/20/26: Added 1 more bilingual over 12 mos old (N = 25)
```{r, include=FALSE}
library(tidyverse)

#Reading in the timecourse output and plotting the bilingual timecourse in R.
btime <- read_csv(
  "bilingual_timecourse.csv",
  col_names = TRUE
)

btime <- btime %>%
  rename(condition = 1)

btime_long <- btime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Bilinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
btime_long <- btime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    ),
    word = str_remove(condition, "^(quiet|noise)_"),
    language = case_when(
      word %in% c("stroller", "hand", "car", "foot") ~ "English",
      word %in% c("perro", "nariz", "leche", "ojo") ~ "Spanish"
    )
  )
```

```{r, include=FALSE}
btime_plot <- btime_long %>%
  group_by(time, quiet_noise, language) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(btime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  facet_wrap(~ language) +
  labs(
    title = "Bilingual Timecourse by Language N=25",
    x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```
## Accuracy Analysis: baseline- corrected (Added 7/20/26)
```{r}
library(tidyverse)

# Reads one window pass. Recomputes proportion from frame counts (cols 5 = Frames to
# Target, 6 = Frames to Distractor) so NaN/DROPPED rows are handled by the count filter,
# not by parsing the accuracy string. total >= 15 frames == 0.5 s of looking at 30 fps.
read_window <- function(path, value) {
  raw <- read.csv(path, header = TRUE, stringsAsFactors = FALSE)
  tibble(
    Participant = as.character(raw[[1]]),
    Trial.Number = raw[[2]],
    Trial.Type   = str_trim(as.character(raw[[4]])),
    ft = suppressWarnings(as.numeric(raw[[5]])),   # Frames to Target
    fd = suppressWarnings(as.numeric(raw[[6]]))    # Frames to Distractor
  ) %>%
    filter(!is.na(Participant), str_trim(Participant) != "") %>%   # drop blank separator rows
    mutate(total = ft + fd) %>%
    filter(!is.na(total), total >= 15) %>%                         # 0.5 s looking; removes DROPPED
    transmute(
      Participant,
      trial = parse_number(as.character(Trial.Number)),            # "Trial 5" -> 5
      Trial.Type,
      !!value := ft / total
    )
}

# Joins a test pass (60:180) with its baseline pass (0:36) and subtracts, per trial.
build_corrected <- function(test_path, base_path, exclude_types) {
  test <- read_window(test_path, "test_acc")
  base <- read_window(base_path, "base_acc")
  inner_join(test, base, by = c("Participant", "trial", "Trial.Type")) %>%   # valid in BOTH windows
    mutate(
      corrected_acc = test_acc - base_acc,
      quiet_noise = case_when(
        str_starts(Trial.Type, "quiet") ~ "Quiet",
        str_starts(Trial.Type, "noise") ~ "Noise"
      )
    ) %>%
    filter(!Trial.Type %in% exclude_types)
}
```

## Bilinguals Accuracy (Updated 7/20/26)

-   This data reflects looking time data for a 4 second window from 2-6 seconds
- baseline-corrected proportion is calculated using baseline from 0-1.2 seconds

### Build corrected data and labels
```{r, include=FALSE}
b_exclude <- c("baseline_1","baseline_2","baseline_3","baseline_4",
               "quiet_eye","noise_eye","noise_nose","quiet_nose",
               "noise_milk","noise_dog","quiet_dog","quiet_milk")

bcorr <- build_corrected("bwindow.csv", "bbaseline.csv", b_exclude) %>%
  mutate(
    word = str_remove(Trial.Type, "^(quiet|noise)_"),
    language = case_when(
      word %in% c("stroller","hand","car","foot") ~ "English",
      word %in% c("perro","nariz","leche","ojo")  ~ "Spanish"
    )
  )
```

### check how many trials survive the baseline requirement (run once, then decide):

```{r}
# The baseline window is only 36 frames, so requiring 15 (=0.5 s) is a HIGH bar there.
# See how much you lose to "no valid baseline" before trusting the corrected numbers.
b_test_n <- read_window("bwindow.csv", "test_acc") %>%
  filter(!Trial.Type %in% b_exclude) %>% nrow()
b_corr_n <- nrow(bcorr)
message("Bilingual trials with valid TEST window: ", b_test_n,
        " | with valid test AND baseline: ", b_corr_n,
        " (", round(100 * b_corr_n / b_test_n), "% retained)")

```
### inclusion rules

```{r}
MIN_PER_COND <- 4   # >=4 valid corrected Quiet AND >=4 valid corrected Noise trials

keep_ids <- bcorr %>%
  count(Participant, quiet_noise) %>%
  pivot_wider(names_from = quiet_noise, values_from = n, values_fill = 0) %>%
  filter(Quiet >= MIN_PER_COND, Noise >= MIN_PER_COND) %>%
  pull(Participant)

bcorr_clean <- bcorr %>% filter(Participant %in% keep_ids)
message("Kept ", length(keep_ids), " of ", n_distinct(bcorr$Participant), " participants")
setdiff(unique(bcorr$Participant), keep_ids)   # who was dropped & why
```
### participant-level means

```{r}
p_level <- bcorr_clean %>%
  group_by(Participant, quiet_noise, language) %>%
  summarise(acc = mean(corrected_acc), n = n(), .groups = "drop")

p_level %>%
  group_by(quiet_noise, language) %>%
  summarise(mean_corr = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")
```
### tests against 0 (the null is now "no shift from baseline"):

```{r}
p_overall <- bcorr_clean %>%
  group_by(Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop")
t.test(p_overall$acc, mu = 0)          # is target looking above pre-naming baseline?

by_cond_wide <- bcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)
t.test(by_cond_wide$Quiet, by_cond_wide$Noise, paired = TRUE)   # Quiet vs Noise
```
### split by language

```{r}
# Bilingual: language x condition, participant-level means, one-sample t vs 0 per cell.
bilingual_lang_cond <- bcorr_clean %>%
  group_by(language, quiet_noise, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(language, quiet_noise) %>%
  summarise(
    N       = n(),
    Mean    = mean(acc),
    SE      = sd(acc) / sqrt(n()),
    CI_low  = t.test(acc, mu = 0)$conf.int[1],
    CI_high = t.test(acc, mu = 0)$conf.int[2],
    t       = t.test(acc, mu = 0)$statistic,
    df      = t.test(acc, mu = 0)$parameter,
    p       = t.test(acc, mu = 0)$p.value,
    .groups = "drop"
  ) %>%
  mutate(across(c(Mean, SE, CI_low, CI_high, t), ~round(., 3)),
         p = round(p, 4)) %>%
  arrange(language, quiet_noise)

bilingual_lang_cond
```

### mixed effects model

```{r}
library(lme4); library(lmerTest)
m <- lmer(corrected_acc ~ quiet_noise * language + (1 | Participant), data = bcorr_clean)
summary(m)
```

### Ignore for now
Plotting accuracy by participant and condition
```{r}
library(dplyr)
library(ggplot2)

# Calculate mean accuracy for each participant in each condition
acc_b <- imputed_b %>%
  group_by(Participant, quiet_noise) %>%
  summarize(
    Accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(acc_b,
       aes(x = quiet_noise,
           y = Accuracy,
           group = Participant)) +
  geom_line(alpha = .4) +
  geom_point(size = 2) +
  theme_classic() +
  labs(
    x = "Condition",
    y = "Mean Accuracy"
  )
```

Calculating a noise difference score (quiet - noise)
```{r}
diff_b <- acc_b %>%
  tidyr::pivot_wider(
    names_from = quiet_noise,
    values_from = Accuracy
  ) %>%
  mutate(
    NoiseDrop = Quiet - Noise
  ) %>%
  arrange(desc(NoiseDrop))

head(diff_b)
```

Plotting diff scores so we can see:
- participants who perform much worse in noise (large positive values),
- participants with little effect of noise (near 0),
- participants who oddly perform better in noise (negative values).

```{r}
ggplot(diff_b,
       aes(x = reorder(Participant, NoiseDrop),
           y = NoiseDrop)) +
  geom_point() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Participant",
    y = "Quiet - Noise Accuracy"
  )
```

## Monolinguals Timecourse (Updated 7/20/26)

There are 16 different trial types for monlinguals:

-   quiet (8): stroller, hand, car, foot, eye, nose, dog, milk

-   noise (8): hand, foot, stroller, car, eye, nose, milk, dog

Object pairs:

- eye/dog
- car/foot
- stroller/hand
- milk/nose

Notes from analysis 6/17/26: Only including monolinguals over 12 months old (N=28)

Notes from analysis 7/20/26: Added 2 more monolinguals over 12 mos (N = 30)
```{r, include=FALSE}
mtime <- read_csv(
  "monolingual_timecourse.csv",
  col_names = TRUE
)

mtime <- mtime %>%
  rename(condition = 1)

mtime_long <- mtime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Monolinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
mtime_long <- mtime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    )
  )
```

```{r, include=FALSE}
mtime_plot <- mtime_long %>%
  group_by(time, quiet_noise) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(mtime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
   title = "Monolingual Timecourse N=30",
   x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

## Monolinguals Accuracy (Updated 7/20/26)

### build corrected data
```{r}
# Monolinguals hear all 8 objects in English; drop only the baseline trials.
m_exclude <- c("baseline_1", "baseline_2", "baseline_3", "baseline_4")

mcorr <- build_corrected("mwindow.csv", "mbaseline.csv", m_exclude)
```

### baseline survival

```{r}
m_test_n <- read_window("mwindow.csv", "test_acc") %>%
  filter(!Trial.Type %in% m_exclude) %>% nrow()
m_corr_n <- nrow(mcorr)
message("Monolingual trials with valid TEST window: ", m_test_n,
        " | with valid test AND baseline: ", m_corr_n,
        " (", round(100 * m_corr_n / m_test_n), "% retained)")
```
### inclusion rules

```{r}
m_keep_ids <- mcorr %>%
  count(Participant, quiet_noise) %>%
  pivot_wider(names_from = quiet_noise, values_from = n, values_fill = 0) %>%
  filter(Quiet >= MIN_PER_COND, Noise >= MIN_PER_COND) %>%
  pull(Participant)

mcorr_clean <- mcorr %>% filter(Participant %in% m_keep_ids)
message("Kept ", length(m_keep_ids), " of ", n_distinct(mcorr$Participant), " participants")
setdiff(unique(mcorr$Participant), m_keep_ids)   # who was dropped & why
```
### participant level means

```{r}
m_p_level <- mcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), n = n(), .groups = "drop")

m_p_level %>%
  group_by(quiet_noise) %>%
  summarise(mean_corr = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")
```
### tests against zero

```{r}
m_overall <- mcorr_clean %>%
  group_by(Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop")
t.test(m_overall$acc, mu = 0)          # is target looking above pre-naming baseline?

m_by_cond_wide <- mcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)
t.test(m_by_cond_wide$Quiet, m_by_cond_wide$Noise, paired = TRUE)   # Quiet vs Noise
```

### mixed model

```{r}
library(lme4); library(lmerTest)
mm <- lmer(corrected_acc ~ quiet_noise + (1 | Participant), data = mcorr_clean)
summary(mm)
```
## Combined summary table

```{r}
library(broom)

# Helper: one-sample t vs 0 on participant-level corrected means, returned as a tidy row
summ_vs0 <- function(df, group_label) {
  p <- df %>%
    group_by(Participant) %>%
    summarise(acc = mean(corrected_acc), .groups = "drop")
  tt <- t.test(p$acc, mu = 0)
  tibble(
    Group   = group_label,
    N       = nrow(p),
    Mean    = unname(tt$estimate),
    CI_low  = tt$conf.int[1],
    CI_high = tt$conf.int[2],
    t       = unname(tt$statistic),
    df      = unname(tt$parameter),
    p       = tt$p.value
  )
}

summary_table <- bind_rows(
  summ_vs0(mcorr_clean, "Monolingual"),
  summ_vs0(bcorr_clean, "Bilingual")
) %>%
  mutate(across(c(Mean, CI_low, CI_high, t), ~round(., 3)),
         p = round(p, 4))

summary_table
```
### Ignore for now
Plotting accuracy by participant and condition
```{r}
# Calculate mean accuracy for each participant in each condition
acc_m <- imputed_m %>%
  group_by(Participant, quiet_noise) %>%
  summarize(
    Accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(acc_m,
       aes(x = quiet_noise,
           y = Accuracy,
           group = Participant)) +
  geom_line(alpha = .4) +
  geom_point(size = 2) +
  theme_classic() +
  labs(
    x = "Condition",
    y = "Mean Accuracy"
  )

```

Calculating a noise difference score (quiet - noise)
```{r}
diff_m <- acc_m %>%
  tidyr::pivot_wider(
    names_from = quiet_noise,
    values_from = Accuracy
  ) %>%
  mutate(
    NoiseDrop = Quiet - Noise
  ) %>%
  arrange(desc(NoiseDrop))

head(diff_m)
```

Plotting diff scores so we can see:
- participants who perform much worse in noise (large positive values),
- participants with little effect of noise (near 0),
- participants who oddly perform better in noise (negative values).

```{r}
ggplot(diff_m,
       aes(x = reorder(Participant, NoiseDrop),
           y = NoiseDrop)) +
  geom_point() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Participant",
    y = "Quiet - Noise Accuracy"
  )
```

# Habituation Task
## Data Cleaning (updated 6/16/26)
- 79 participants (including V1 and V2 data points) with habituation data
this only includes kids who did habituate
- need to filter through kids and pick one data point (either V1 or V2) for every participant

```{r}
library(dplyr)
hab <- read.csv("hab.csv")
age <- read.csv("age.csv")
lena <- read.csv("lena.csv")
mcdi <- read.csv("mcdi.csv")

hab <- hab %>%
  left_join(age, by = "Participant")

hab <- hab %>%
  left_join(lena, by = "Participant")

hab <- hab %>%
  left_join(mcdi, by = "Participant")

# add a language column
hab <- hab %>%
  mutate(
    language = case_when(
      grepl("^M", Participant) ~ "monolingual",
      grepl("^B", Participant) ~ "bilingual",
      TRUE ~ NA_character_
    )
  )
```

### Missing Data
```{r}
#which kids are missing lena data
hab %>%
  filter(is.na(noise_sum)) %>%
  pull(Participant)

#19 kids 
#[1] "M05ES"     "M08SA"     "M43EW"     "M47JJ"     "M48DZ"    
#[6] "M08SA_V2"  "M24JB_V2"  "M37JD_V2"  "B04DR"     "B06JR"    
#[11] "B10KG"     "B17YT"     "B22JR"     "B24EBR"    "B23RF"    
#[16] "B29NRP"    "B31ZR"     "B32NF"     "B12KRM_V2"
```

```{r}
#which kids are missing mcdi data
hab %>%
  filter(is.na(mcdi)) %>%
  pull(Participant)

#32 kids 
#[1] "M11BLP"    "M22AS"     "M41GG"     "M43EW"     "M45DA"    
#[6] "M47JJ"     "M48DZ"     "M46CC"     "M12RO_V2"  "M13CY_V2" 
#[11] "M37JD_V2"  "B05ET"     "B10KG"     "B16LR"     "B17YT"    
#[16] "B18LS"     "B21HM"     "B22JR"     "B24EBR"    "B23RF"    
#[21] "B27KDS"    "B28APF"    "B30MR"     "B29NRP"    "B31ZR"    
#[26] "B34ES"     "B26GB"     "B32NF"     "B33KLH"    "B05ET_V2" 
#[31] "B11ARF_V2" "B12KRM_V2"
```

```{r}
# seeing how many kids have a V2 data point for habit data 
hab$visit <- ifelse(grepl("_V2$", hab$Participant), "V2", "V1")

table(hab$language, hab$visit)

#            V1 V2
#bilingual   28  3
#monolingual 32 16
```

### Age groups
```{r}
# creating age group bins
hab <- hab %>%
  mutate(
    age_group = case_when(
      age >= 8  & age < 13 ~ "8–12.9",
      age >= 13 & age < 19 ~ "13–18.9",
      age >= 19 & age < 27 ~ "19–26.9",
      TRUE ~ NA_character_
    ),
    age_group = factor(
      age_group,
      levels = c("8–12.9", "13–18.9", "19–26.9")
    )
  )

library(ggplot2)
ggplot(hab, aes(x = age)) +
  geom_histogram(binwidth = 1) +
  geom_rug() +
  theme_classic()
```

## Stats
### Linear regressions
```{r}
library(lme4)

m1 <- lm(nov_score ~ age * language, data = hab)
summary(m1)

m2 <- lm(hab_trials ~ age * language, data = hab)
summary(m2)

m3 <- lm(nov1 ~ age * language, data = hab)
summary(m3)

m4 <- lm(nov_score ~ tv, data = bi)
summary(m4)
# tv main effect p=0.0525 for bilingual sample

m5 <- lm(nov_score ~ noise_sum, data = mono)
summary(m5)

m6 <- lm(hab_trials ~ tv, data = bi)
summary(m6)

m7 <- lm(hab_trials ~ noise_sum, data = mono)
summary(m7)

m8 <- lm(nov1 ~ tv, data = bi)
summary(m8)

m9 <- lm(nov1 ~ noise_sum, data = mono)
summary(m9)
```

## Plotting Novelty Response Score
```{r}
library(ggplot2)

ggplot(hab, aes(x = language, y = nov_score)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  labs(
    x = "Language Group",
    y = "Novelty Response Score",
    title = "Novelty Response Scores by Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab, aes(x = age, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Age (months)",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab, aes(x = age_group, y = nov_score, color = language)) +
  geom_boxplot(position = position_dodge(width = 0.8),
               outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.15,
    dodge.width = 0.8
  ),
  alpha = 0.7) +
  labs(
    x = "Age Group (months)",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

## Hab + LENA plots
```{r}
ggplot(hab, aes(x = noise_sum, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "LENA Total Noise",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab, aes(x = tv, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "LENA TV Noise",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```
