---
title: "BIN_Analysis"
output: html_document
date: "2026-06-01"
---
# Word Recognition Task
## Bilinguals Timecourse (Updated 6/17/26)

There are 16 different trial types for bilinguals:

-   quiet (8): stroller, hand, car, foot, perro, nariz, leche, ojo

-   noise (8): hand, foot, stroller, car, perro, nariz, ojo, leche

Object pairs:

- ojo/perro
- car/foot
- stroller/hand
- nariz/leche

Notes from analysis 6/17/26: Only including bilinguals over 12 months old (N=24)
```{r, include=FALSE}
library(tidyverse)

#Reading in the timecourse output and plotting the bilingual timecourse in R.
btime <- read_csv(
  "bilingual_timecourse.csv",
  col_names = TRUE
)

btime <- btime %>%
  rename(condition = 1)

btime_long <- btime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Bilinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
btime_long <- btime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    )
  )
```

```{r, include=FALSE}
btime_plot <- btime_long %>%
  group_by(time, quiet_noise) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(btime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
   title = "Bilingual Timecourse N=24",
   x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

## Bilinguals Accuracy (Updated 6/18/26)

-   This data reflects looking time data for a 4.2 second window from 2 seconds until the end of the trial (6.2 seconds)
- This window was decided mostly based on the monolingual timecourse plot below 

```{r, include=FALSE}
bwindow <- read.csv("bwindow.csv", header=TRUE,na.strings=c(""))

#remove rows based on NA in Participant column (empty rows)
bwindow <- bwindow %>% drop_na(Participant)

#remove trials that we don't want based on the name
bwindow <- bwindow[!bwindow$Trial.Type %in% c("baseline_1", "baseline_2", "baseline_3", "baseline_4","quiet_eye", "noise_eye", "noise_nose", "quiet_nose", "noise_milk", "noise_dog", "quiet_dog", "quiet_milk"),]

bwindow <- bwindow %>%
  mutate(
    quiet_noise = case_when(
      str_starts(Trial.Type, "quiet") ~ "Quiet",
      str_starts(Trial.Type, "noise") ~ "Noise"
    )
  )

rownames(bwindow) <- NULL
```

```{r, warning=FALSE}
bwindow %>%
  group_by(quiet_noise) %>%
  summarise(
    mean_accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

# quiet_noise mean_accuracy
# Noise               0.528
# Quiet               0.521
```

```{r}
#imputing data -> impute NaN values with the means of the type accuracy for that trial type (using simplified quiet_noise variable)
imputed_b <- bwindow %>%
  group_by(quiet_noise) %>%
  mutate(Type.Accuracy = ifelse(is.na(Type.Accuracy), mean(Type.Accuracy, na.rm = TRUE), Type.Accuracy))
```

```{r}
# make z scores in order to check for outliers
imputed_b$z <- ave(imputed_b$Type.Accuracy, imputed_b$quiet_noise, FUN = function(x) (x - mean(x)) / sd(x))

#check for outliers
outliers_b <- imputed_b[abs(imputed_b$z) > 1.5, ]

#changed z score cut off to 1.5 to get rid of trials with accuracy of 0
#now there are 93 outlier trials 
# there were 0 outlier trials when z score cut off was set at 2
```

```{r}
#remove outliers
imputed_b <- imputed_b[!paste(imputed_b$Participant, imputed_b$quiet_noise, imputed_b$Type.Accuracy) %in% paste(outliers_b$Participant, outliers_b$quiet_noise, outliers_b$Type.Accuracy), ]
```

```{r}
imputed_b %>%
  group_by(quiet_noise) %>%
  summarise(
    mean_accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

#  quiet_noise mean_accuracy
# Noise               0.529
# Quiet               0.558
```

Plotting accuracy by participant and condition
```{r}
library(dplyr)
library(ggplot2)

# Calculate mean accuracy for each participant in each condition
acc_b <- imputed_b %>%
  group_by(Participant, quiet_noise) %>%
  summarize(
    Accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(acc_b,
       aes(x = quiet_noise,
           y = Accuracy,
           group = Participant)) +
  geom_line(alpha = .4) +
  geom_point(size = 2) +
  theme_classic() +
  labs(
    x = "Condition",
    y = "Mean Accuracy"
  )
```

Calculating a noise difference score (quiet - noise)
```{r}
diff_b <- acc_b %>%
  tidyr::pivot_wider(
    names_from = quiet_noise,
    values_from = Accuracy
  ) %>%
  mutate(
    NoiseDrop = Quiet - Noise
  ) %>%
  arrange(desc(NoiseDrop))

head(diff_b)
```

Plotting diff scores so we can see:
- participants who perform much worse in noise (large positive values),
- participants with little effect of noise (near 0),
- participants who oddly perform better in noise (negative values).

```{r}
ggplot(diff_b,
       aes(x = reorder(Participant, NoiseDrop),
           y = NoiseDrop)) +
  geom_point() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Participant",
    y = "Quiet - Noise Accuracy"
  )
```

## Monolinguals Timecourse (Updated 6/17/26)

There are 16 different trial types for monlinguals:

-   quiet (8): stroller, hand, car, foot, eye, nose, dog, milk

-   noise (8): hand, foot, stroller, car, eye, nose, milk, dog

Object pairs:

- eye/dog
- car/foot
- stroller/hand
- milk/nose

Notes from analysis 6/17/26: Only including monolinguals over 12 months old (N=28)
```{r, include=FALSE}
mtime <- read_csv(
  "monolingual_timecourse.csv",
  col_names = TRUE
)

mtime <- mtime %>%
  rename(condition = 1)

mtime_long <- mtime %>%
  pivot_longer(
    cols = -condition,
    names_to = "time",
    values_to = "looking_time"
  ) %>%
  mutate(
    time = as.numeric(time)
  )
```

Monolinguals Quiet v. Noise plot. The dashed line at 1.2 seconds indicates the time at which the target word onset occurred.

```{r, include=FALSE}
#Reformatting the data so we can look at timecourse across quiet and noise conditions only.
mtime_long <- mtime_long %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    )
  )
```

```{r, include=FALSE}
mtime_plot <- mtime_long %>%
  group_by(time, quiet_noise) %>%
  summarise(
    mean_looking = mean(looking_time, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE}
ggplot(mtime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = 1.2,
    linetype = "dashed",
    linewidth = 0.6,
    color = "gray40"
  ) +
  labs(
   title = "Monolingual Timecourse N=28",
   x = "Time (s)",
    y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

## Monolinguals Accuracy (Updated 6/18/26)
- Same 4.2 second window for monolinguals

```{r}
mwindow <- read.csv("mwindow.csv", header=TRUE,na.strings=c(""))

#remove rows based on NA in Participant column (empty rows)
mwindow <- mwindow %>% drop_na(Participant)

#remove trials that we don't want based on the name
mwindow <- mwindow[!mwindow$Trial.Type %in% c("baseline_1", "baseline_2", "baseline_3", "baseline_4", "quiet_ojo", "noise_perro", "noise_nariz", "quiet_perro","quiet_nariz","noise_ojo", "quiet_leche", "noise_leche"),]

mwindow <- mwindow %>%
  mutate(
    quiet_noise = case_when(
      str_starts(Trial.Type, "quiet") ~ "Quiet",
      str_starts(Trial.Type, "noise") ~ "Noise"
    )
  )

rownames(mwindow) <- NULL
```

```{r}
mwindow %>%
  group_by(quiet_noise) %>%
  summarise(
    mean_accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

#  quiet_noise mean_accuracy
# Noise               0.587
# Quiet               0.545
```
```{r}
#imputing data -> impute NaN values with the means of the type accuracy for that trial type (using simplified quiet_noise variable)
imputed_m <- mwindow %>%
  group_by(quiet_noise) %>%
  mutate(Type.Accuracy = ifelse(is.na(Type.Accuracy), mean(Type.Accuracy, na.rm = TRUE), Type.Accuracy))
```

```{r}
# make z scores in order to check for outliers
imputed_m$z <- ave(imputed_m$Type.Accuracy, imputed_m$quiet_noise, FUN = function(x) (x - mean(x)) / sd(x))

#check for outliers
outliers_m <- imputed_m[abs(imputed_m$z) > 1.5, ]

#there were 85 outlier trials for the monolingual group when i switched the z score cutoff to 1.5 instead of 2
```

```{r}
#remove outliers
imputed_m <- imputed_m[!paste(imputed_m$Participant, imputed_m$quiet_noise, imputed_m$Type.Accuracy) %in% paste(outliers_m$Participant, outliers_m$quiet_noise, outliers_m$Type.Accuracy), ]
```

```{r}
imputed_m %>%
  group_by(quiet_noise) %>%
  summarise(
    mean_accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

#  quiet_noise mean_accuracy
# Noise               0.592
# Quiet               0.563
```

Plotting accuracy by participant and condition
```{r}
# Calculate mean accuracy for each participant in each condition
acc_m <- imputed_m %>%
  group_by(Participant, quiet_noise) %>%
  summarize(
    Accuracy = mean(Type.Accuracy, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(acc_m,
       aes(x = quiet_noise,
           y = Accuracy,
           group = Participant)) +
  geom_line(alpha = .4) +
  geom_point(size = 2) +
  theme_classic() +
  labs(
    x = "Condition",
    y = "Mean Accuracy"
  )

```

Calculating a noise difference score (quiet - noise)
```{r}
diff_m <- acc_m %>%
  tidyr::pivot_wider(
    names_from = quiet_noise,
    values_from = Accuracy
  ) %>%
  mutate(
    NoiseDrop = Quiet - Noise
  ) %>%
  arrange(desc(NoiseDrop))

head(diff_m)
```

Plotting diff scores so we can see:
- participants who perform much worse in noise (large positive values),
- participants with little effect of noise (near 0),
- participants who oddly perform better in noise (negative values).

```{r}
ggplot(diff_m,
       aes(x = reorder(Participant, NoiseDrop),
           y = NoiseDrop)) +
  geom_point() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Participant",
    y = "Quiet - Noise Accuracy"
  )
```

# Habituation Task
## Data Cleaning (updated 6/16/26)
- 79 participants (including V1 and V2 data points) with habituation data
this only includes kids who did habituate
- need to filter through kids and pick one data point (either V1 or V2) for every participant

```{r}
library(dplyr)
hab <- read.csv("hab.csv")
age <- read.csv("age.csv")
lena <- read.csv("lena.csv")
mcdi <- read.csv("mcdi.csv")

hab <- hab %>%
  left_join(age, by = "Participant")

hab <- hab %>%
  left_join(lena, by = "Participant")

hab <- hab %>%
  left_join(mcdi, by = "Participant")

# add a language column
hab <- hab %>%
  mutate(
    language = case_when(
      grepl("^M", Participant) ~ "monolingual",
      grepl("^B", Participant) ~ "bilingual",
      TRUE ~ NA_character_
    )
  )
```

### Missing Data
```{r}
#which kids are missing lena data
hab %>%
  filter(is.na(noise_sum)) %>%
  pull(Participant)

#19 kids 
#[1] "M05ES"     "M08SA"     "M43EW"     "M47JJ"     "M48DZ"    
#[6] "M08SA_V2"  "M24JB_V2"  "M37JD_V2"  "B04DR"     "B06JR"    
#[11] "B10KG"     "B17YT"     "B22JR"     "B24EBR"    "B23RF"    
#[16] "B29NRP"    "B31ZR"     "B32NF"     "B12KRM_V2"
```

```{r}
#which kids are missing mcdi data
hab %>%
  filter(is.na(mcdi)) %>%
  pull(Participant)

#32 kids 
#[1] "M11BLP"    "M22AS"     "M41GG"     "M43EW"     "M45DA"    
#[6] "M47JJ"     "M48DZ"     "M46CC"     "M12RO_V2"  "M13CY_V2" 
#[11] "M37JD_V2"  "B05ET"     "B10KG"     "B16LR"     "B17YT"    
#[16] "B18LS"     "B21HM"     "B22JR"     "B24EBR"    "B23RF"    
#[21] "B27KDS"    "B28APF"    "B30MR"     "B29NRP"    "B31ZR"    
#[26] "B34ES"     "B26GB"     "B32NF"     "B33KLH"    "B05ET_V2" 
#[31] "B11ARF_V2" "B12KRM_V2"
```

```{r}
# seeing how many kids have a V2 data point for habit data 
hab$visit <- ifelse(grepl("_V2$", hab$Participant), "V2", "V1")

table(hab$language, hab$visit)

#            V1 V2
#bilingual   28  3
#monolingual 32 16
```

### Age groups
```{r}
# creating age group bins
hab <- hab %>%
  mutate(
    age_group = case_when(
      age >= 8  & age < 13 ~ "8–12.9",
      age >= 13 & age < 19 ~ "13–18.9",
      age >= 19 & age < 27 ~ "19–26.9",
      TRUE ~ NA_character_
    ),
    age_group = factor(
      age_group,
      levels = c("8–12.9", "13–18.9", "19–26.9")
    )
  )

library(ggplot2)
ggplot(hab, aes(x = age)) +
  geom_histogram(binwidth = 1) +
  geom_rug() +
  theme_classic()
```

## Stats
### Linear regressions
```{r}
library(lme4)

m1 <- lm(nov_score ~ age * language, data = hab)
summary(m1)

m2 <- lm(hab_trials ~ age * language, data = hab)
summary(m2)

m3 <- lm(nov1 ~ age * language, data = hab)
summary(m3)

m4 <- lm(nov_score ~ tv, data = bi)
summary(m4)
# tv main effect p=0.0525 for bilingual sample

m5 <- lm(nov_score ~ noise_sum, data = mono)
summary(m5)

m6 <- lm(hab_trials ~ tv, data = bi)
summary(m6)

m7 <- lm(hab_trials ~ noise_sum, data = mono)
summary(m7)

m8 <- lm(nov1 ~ tv, data = bi)
summary(m8)

m9 <- lm(nov1 ~ noise_sum, data = mono)
summary(m9)
```

## Plotting Novelty Response Score
```{r}
library(ggplot2)

ggplot(hab, aes(x = language, y = nov_score)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  labs(
    x = "Language Group",
    y = "Novelty Response Score",
    title = "Novelty Response Scores by Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab, aes(x = age, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Age (months)",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab, aes(x = age_group, y = nov_score, color = language)) +
  geom_boxplot(position = position_dodge(width = 0.8),
               outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.15,
    dodge.width = 0.8
  ),
  alpha = 0.7) +
  labs(
    x = "Age Group (months)",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

## Hab + LENA plots
```{r}
ggplot(hab, aes(x = noise_sum, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "LENA Total Noise",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```

```{r}
ggplot(hab, aes(x = tv, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "LENA TV Noise",
    y = "Novelty Response Score",
    color = "Language Group"
  ) +
  theme_classic()
```
