---
title: "Bilingual Infant Noise (BIN) Study"
subtitle: "Word Recognition & Habituation — Results Walkthrough"
author: "Mels Ayala"
date: today

format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: false
    pdf-engine: pdflatex
    mainfont: "Latin Modern Roman"
    include-in-header: |
      \addtokomafont{disposition}{\rmfamily}
    geometry: margin=1in

execute:
  echo: false
  warning: false
  message: false

editor: source
---

```{r setup, include=FALSE}
# One place for every package + global options. Hidden from the PDF.
# Required packages: tidyverse, lme4, lmerTest, broom, emmeans, car
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(emmeans)

knitr::opts_chunk$set(
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 150,
  fig.align  = "center"
)
options(width = 85)   # keeps wide model summaries from overflowing the page
```

```{r ses-prep, include=FALSE}
# --- SES predictors ---
# Bracket -> ordinal rank (1 = lowest). With ">$75,000" open-ended, ranks are
# more defensible than midpoints. parent_ed is z-scored across the full sample.
# clean_id() normalizes IDs (trim + uppercase) so the SES join below doesn't
# silently drop participants to whitespace/case mismatches.
clean_id <- function(x) str_trim(toupper(as.character(x)))

income_levels <- c("<$15,000", "$15,000-$25,000", "$25,000-$35,000",
                   "$35,000-$45,000", "$45,000-$55,000", "$55,000-$65,000",
                   "$65,000-$75,000", ">$75,000")

ses <- read_csv("predictors.csv", show_col_types = FALSE) %>%
  select(participant, `parent ed`, income) %>%
  rename(Participant = participant, parent_ed = `parent ed`) %>%
  mutate(
    Participant = clean_id(Participant),
    income_rank = as.integer(factor(str_trim(income), levels = income_levels)),
    parent_ed_z = as.numeric(scale(parent_ed))
  )

# expected missing SES: M43EW, B17YT, B23RF, B29NRP, B31ZR
```

```{r wordrec-functions, include=FALSE}
# Reads one window pass. Recomputes proportion from frame counts (col 5 = Frames
# to Target, 6 = Frames to Distractor) so NaN/DROPPED rows are handled by the
# count filter. total >= 15 frames == 0.5 s of looking at 30 fps.
read_window <- function(path, value) {
  raw <- read.csv(path, header = TRUE, stringsAsFactors = FALSE)
  tibble(
    Participant  = as.character(raw[[1]]),
    Trial.Number = raw[[2]],
    Trial.Type   = str_trim(as.character(raw[[4]])),
    ft = suppressWarnings(as.numeric(raw[[5]])),
    fd = suppressWarnings(as.numeric(raw[[6]]))
  ) %>%
    filter(!is.na(Participant), str_trim(Participant) != "") %>%
    mutate(total = ft + fd) %>%
    filter(!is.na(total), total >= 15) %>%
    transmute(
      Participant,
      trial = parse_number(as.character(Trial.Number)),
      Trial.Type,
      !!value := ft / total
    )
}

# Joins a test pass (60:180) with its baseline pass (0:36) and subtracts, per trial.
build_corrected <- function(test_path, base_path, exclude_types) {
  test <- read_window(test_path, "test_acc")
  base <- read_window(base_path, "base_acc")
  inner_join(test, base, by = c("Participant", "trial", "Trial.Type")) %>%
    mutate(
      corrected_acc = test_acc - base_acc,
      quiet_noise = case_when(
        str_starts(Trial.Type, "quiet") ~ "Quiet",
        str_starts(Trial.Type, "noise") ~ "Noise"
      )
    ) %>%
    filter(!Trial.Type %in% exclude_types)
}
```

# Word Recognition Task

## Bilingual Timecourse

There are 16 trial types for bilinguals:

- quiet (8): stroller, hand, car, foot, perro, nariz, leche, ojo
- noise (8): hand, foot, stroller, car, perro, nariz, ojo, leche

Object pairs: ojo/perro, car/foot, stroller/hand, nariz/leche.

Includes bilinguals over 12 months old (N = 25, updated 7/20/26). The dashed
line at 1.2 s marks target word onset.

```{r btime-read, include=FALSE}
btime <- read_csv("bilingual_timecourse.csv", col_names = TRUE) %>%
  rename(condition = 1)

btime_long <- btime %>%
  pivot_longer(cols = -condition, names_to = "time", values_to = "looking_time") %>%
  mutate(time = as.numeric(time)) %>%
  mutate(
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    ),
    word = str_remove(condition, "^(quiet|noise)_"),
    language = case_when(
      word %in% c("stroller", "hand", "car", "foot") ~ "English",
      word %in% c("perro", "nariz", "leche", "ojo")  ~ "Spanish"
    )
  )

btime_plot <- btime_long %>%
  group_by(time, quiet_noise, language) %>%
  summarise(mean_looking = mean(looking_time, na.rm = TRUE), .groups = "drop")
```

```{r btime-plot}
ggplot(btime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 1.2, linetype = "dashed", linewidth = 0.6, color = "gray40") +
  facet_wrap(~ language) +
  labs(
    title = "Bilingual Timecourse by Language (N = 25)",
    x = "Time (s)", y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

## Bilingual Accuracy (baseline-corrected)

Looking-time data for a 4 s window (2–6 s); baseline-corrected using the
0–1.2 s window.

```{r bcorr-build, include=FALSE}
b_exclude <- c("baseline_1","baseline_2","baseline_3","baseline_4",
               "quiet_eye","noise_eye","noise_nose","quiet_nose",
               "noise_milk","noise_dog","quiet_dog","quiet_milk")

bcorr <- build_corrected("bwindow.csv", "bbaseline.csv", b_exclude) %>%
  mutate(
    word = str_remove(Trial.Type, "^(quiet|noise)_"),
    language = case_when(
      word %in% c("stroller","hand","car","foot") ~ "English",
      word %in% c("perro","nariz","leche","ojo")  ~ "Spanish"
    )
  )
```

**Baseline survival.** The baseline window is only 36 frames, so requiring
15 (= 0.5 s) is a high bar. How many trials survive before we trust the
corrected numbers:

```{r bcorr-survival}
b_test_n <- read_window("bwindow.csv", "test_acc") %>%
  filter(!Trial.Type %in% b_exclude) %>% nrow()
b_corr_n <- nrow(bcorr)
cat("Bilingual trials with valid TEST window:", b_test_n,
    "| with valid test AND baseline:", b_corr_n,
    paste0("(", round(100 * b_corr_n / b_test_n), "% retained)\n"))
```

**Inclusion rule:** keep participants with >= 4 valid corrected Quiet *and*
>= 4 valid corrected Noise trials.

```{r bcorr-inclusion}
MIN_PER_COND <- 4

keep_ids <- bcorr %>%
  count(Participant, quiet_noise) %>%
  pivot_wider(names_from = quiet_noise, values_from = n, values_fill = 0) %>%
  filter(Quiet >= MIN_PER_COND, Noise >= MIN_PER_COND) %>%
  pull(Participant)

bcorr_clean <- bcorr %>% filter(Participant %in% keep_ids)
cat("Kept", length(keep_ids), "of", n_distinct(bcorr$Participant), "participants\n")
```

**SES join.** IDs are normalized on both sides before joining. The check
below should return an empty tibble — if it lists anyone, those looking-data
IDs still don't match a row in `predictors.csv`.

```{r bcorr-ses-join}
bcorr_ses <- bcorr_clean %>%
  mutate(Participant = clean_id(Participant)) %>%
  left_join(ses, by = "Participant")

bcorr_ses %>% filter(is.na(parent_ed)) %>% distinct(Participant)
```

### Participant-level means

```{r bcorr-pmeans}
p_level <- bcorr_clean %>%
  group_by(Participant, quiet_noise, language) %>%
  summarise(acc = mean(corrected_acc), n = n(), .groups = "drop")

p_level %>%
  group_by(quiet_noise, language) %>%
  summarise(mean_corr = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")
```

### Tests against 0 (null = no shift from baseline)

```{r bcorr-ttests}
p_overall <- bcorr_clean %>%
  group_by(Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop")
t.test(p_overall$acc, mu = 0)          # target looking above pre-naming baseline?

by_cond_wide <- bcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)
t.test(by_cond_wide$Quiet, by_cond_wide$Noise, paired = TRUE)   # Quiet vs Noise
```

### Split by language

```{r bcorr-by-language}
bilingual_lang_cond <- bcorr_clean %>%
  group_by(language, quiet_noise, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(language, quiet_noise) %>%
  summarise(
    N       = n(),
    Mean    = mean(acc),
    SE      = sd(acc) / sqrt(n()),
    CI_low  = t.test(acc, mu = 0)$conf.int[1],
    CI_high = t.test(acc, mu = 0)$conf.int[2],
    t       = t.test(acc, mu = 0)$statistic,
    df      = t.test(acc, mu = 0)$parameter,
    p       = t.test(acc, mu = 0)$p.value,
    .groups = "drop"
  ) %>%
  mutate(across(c(Mean, SE, CI_low, CI_high, t), ~round(., 3)),
         p = round(p, 4)) %>%
  arrange(language, quiet_noise)

bilingual_lang_cond
```

### Mixed-effects model, with vs. without SES

Both models are fit on the SES-complete subset so the comparison is
apples-to-apples. `anova()` tests whether adding parent education improves fit.

```{r bilingual-mixed-ses}
b_ses_complete <- bcorr_ses %>% filter(!is.na(parent_ed_z))

m0 <- lmer(corrected_acc ~ quiet_noise * language + (1 | Participant),
           data = b_ses_complete)
m1 <- lmer(corrected_acc ~ quiet_noise * language + parent_ed_z + (1 | Participant),
           data = b_ses_complete)

summary(m1)
anova(m0, m1)
cat("N participants in SES models:", n_distinct(b_ses_complete$Participant), "\n")
```

For reference, the same model on the **full** bilingual sample (no SES filter):

```{r bilingual-mixed-full}
m <- lmer(corrected_acc ~ quiet_noise * language + (1 | Participant), data = bcorr_clean)
summary(m)
```

### Accuracy by object

```{r bcorr-by-object}
b_by_object <- bcorr_clean %>%
  group_by(language, word, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(language, word) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

ggplot(b_by_object, aes(x = mean, y = reorder(word, mean), color = language)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_pointrange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.8) +
  labs(
    title = "Bilingual: Baseline-Corrected Accuracy by Word",
    x = "Corrected accuracy (target looking above baseline)",
    y = NULL, color = "Language"
  ) +
  theme_classic()
```

### Recognized words (exploratory)

Words selected post-hoc for being above 0.

```{r bcorr-goodwords}
good_words <- c("leche", "car", "perro", "hand")
b_good <- bcorr_clean %>% filter(word %in% good_words)

b_good_cell <- b_good %>%
  group_by(word, quiet_noise, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(word, quiet_noise) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

b_good_cell
```

```{r bcorr-goodwords-plot}
ggplot(b_good_cell, aes(x = quiet_noise, y = mean, fill = quiet_noise)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  geom_hline(yintercept = 0, color = "gray40") +
  facet_wrap(~ word) +
  labs(
    title = "Bilingual: Quiet vs. Noise for Recognized Words",
    subtitle = "Exploratory — words selected post-hoc for being above 0",
    x = NULL, y = "Corrected accuracy", fill = "Condition"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r bcorr-goodwords-ttest}
b_good_wide <- b_good %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)

t.test(b_good_wide$Quiet, b_good_wide$Noise, paired = TRUE)   # exploratory
```

## Monolingual Timecourse

16 trial types for monolinguals:

- quiet (8): stroller, hand, car, foot, eye, nose, dog, milk
- noise (8): hand, foot, stroller, car, eye, nose, milk, dog

Object pairs: eye/dog, car/foot, stroller/hand, milk/nose. Includes
monolinguals over 12 months old (N = 30, updated 7/20/26).

```{r mtime-read, include=FALSE}
mtime <- read_csv("monolingual_timecourse.csv", col_names = TRUE) %>%
  rename(condition = 1)

mtime_plot <- mtime %>%
  pivot_longer(cols = -condition, names_to = "time", values_to = "looking_time") %>%
  mutate(
    time = as.numeric(time),
    quiet_noise = case_when(
      str_starts(condition, "quiet") ~ "Quiet",
      str_starts(condition, "noise") ~ "Noise"
    )
  ) %>%
  group_by(time, quiet_noise) %>%
  summarise(mean_looking = mean(looking_time, na.rm = TRUE), .groups = "drop")
```

```{r mtime-plot}
ggplot(mtime_plot, aes(x = time, y = mean_looking, color = quiet_noise)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 1.2, linetype = "dashed", linewidth = 0.6, color = "gray40") +
  labs(
    title = "Monolingual Timecourse (N = 30)",
    x = "Time (s)", y = "Proportion of Looking Time to Target",
    color = "Listening Condition"
  ) +
  theme_classic()
```

## Monolingual Accuracy (baseline-corrected)

```{r mcorr-build, include=FALSE}
# Monolinguals hear all 8 objects in English; drop only the baseline trials.
m_exclude <- c("baseline_1", "baseline_2", "baseline_3", "baseline_4")
mcorr <- build_corrected("mwindow.csv", "mbaseline.csv", m_exclude)
```

**Baseline survival:**

```{r mcorr-survival}
m_test_n <- read_window("mwindow.csv", "test_acc") %>%
  filter(!Trial.Type %in% m_exclude) %>% nrow()
m_corr_n <- nrow(mcorr)
cat("Monolingual trials with valid TEST window:", m_test_n,
    "| with valid test AND baseline:", m_corr_n,
    paste0("(", round(100 * m_corr_n / m_test_n), "% retained)\n"))
```

**Inclusion rule** (same as bilinguals):

```{r mcorr-inclusion}
m_keep_ids <- mcorr %>%
  count(Participant, quiet_noise) %>%
  pivot_wider(names_from = quiet_noise, values_from = n, values_fill = 0) %>%
  filter(Quiet >= MIN_PER_COND, Noise >= MIN_PER_COND) %>%
  pull(Participant)

mcorr_clean <- mcorr %>% filter(Participant %in% m_keep_ids)
cat("Kept", length(m_keep_ids), "of", n_distinct(mcorr$Participant), "participants\n")
```

```{r mcorr-ses-join}
mcorr_ses <- mcorr_clean %>%
  mutate(Participant = clean_id(Participant)) %>%
  left_join(ses, by = "Participant")

mcorr_ses %>% filter(is.na(parent_ed)) %>% distinct(Participant)   # should be empty
```

### Participant-level means

```{r mcorr-pmeans}
m_p_level <- mcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), n = n(), .groups = "drop")

m_p_level %>%
  group_by(quiet_noise) %>%
  summarise(mean_corr = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")
```

### Tests against zero

```{r mcorr-ttests}
m_overall <- mcorr_clean %>%
  group_by(Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop")
t.test(m_overall$acc, mu = 0)

m_by_cond_wide <- mcorr_clean %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)
t.test(m_by_cond_wide$Quiet, m_by_cond_wide$Noise, paired = TRUE)
```

### Accuracy by object

```{r mcorr-by-object}
m_by_object <- mcorr_clean %>%
  mutate(word = str_remove(Trial.Type, "^(quiet|noise)_")) %>%
  group_by(word, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(word) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

ggplot(m_by_object, aes(x = mean, y = reorder(word, mean))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_pointrange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.8) +
  labs(
    title = "Monolingual: Baseline-Corrected Accuracy by Word",
    x = "Corrected accuracy (target looking above baseline)", y = NULL
  ) +
  theme_classic()
```

### Recognized words (exploratory)

```{r mcorr-goodwords}
m_good_words <- c("milk", "hand", "car", "dog")

m_good <- mcorr_clean %>%
  mutate(word = str_remove(Trial.Type, "^(quiet|noise)_")) %>%
  filter(word %in% m_good_words)

m_good_cell <- m_good %>%
  group_by(word, quiet_noise, Participant) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  group_by(word, quiet_noise) %>%
  summarise(mean = mean(acc), se = sd(acc)/sqrt(n()), n = n(), .groups = "drop")

m_good_cell
```

```{r mcorr-goodwords-plot}
ggplot(m_good_cell, aes(x = quiet_noise, y = mean, fill = quiet_noise)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  geom_hline(yintercept = 0, color = "gray40") +
  facet_wrap(~ word) +
  labs(
    title = "Monolingual: Quiet vs. Noise for Recognized Words",
    subtitle = "Exploratory — words selected post-hoc for being above 0",
    x = NULL, y = "Corrected accuracy", fill = "Condition"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r mcorr-goodwords-ttest}
m_good_wide <- m_good %>%
  group_by(Participant, quiet_noise) %>%
  summarise(acc = mean(corrected_acc), .groups = "drop") %>%
  pivot_wider(names_from = quiet_noise, values_from = acc)

t.test(m_good_wide$Quiet, m_good_wide$Noise, paired = TRUE)   # exploratory
```

### Mixed-effects model, with vs. without SES

Little is expected here: monolingual parent education has a restricted range
(mostly 16–17 years).

```{r mono-mixed-ses}
m_ses_complete <- mcorr_ses %>% filter(!is.na(parent_ed_z))

mm0 <- lmer(corrected_acc ~ quiet_noise + (1 | Participant), data = m_ses_complete)
mm1 <- lmer(corrected_acc ~ quiet_noise + parent_ed_z + (1 | Participant), data = m_ses_complete)

summary(mm1)
anova(mm0, mm1)
cat("N participants in SES models:", n_distinct(m_ses_complete$Participant), "\n")
```

```{r mono-mixed-full}
mm <- lmer(corrected_acc ~ quiet_noise + (1 | Participant), data = mcorr_clean)
summary(mm)
```

## Combined Analyses

### Summary table (both groups vs. 0)

```{r combined-summary}
summ_vs0 <- function(df, group_label) {
  p <- df %>% group_by(Participant) %>% summarise(acc = mean(corrected_acc), .groups = "drop")
  tt <- t.test(p$acc, mu = 0)
  tibble(
    Group = group_label, N = nrow(p),
    Mean = unname(tt$estimate),
    CI_low = tt$conf.int[1], CI_high = tt$conf.int[2],
    t = unname(tt$statistic), df = unname(tt$parameter), p = tt$p.value
  )
}

summary_table <- bind_rows(
  summ_vs0(mcorr_clean, "Monolingual"),
  summ_vs0(bcorr_clean, "Bilingual")
) %>%
  mutate(across(c(Mean, CI_low, CI_high, t), ~round(., 3)), p = round(p, 4))

summary_table
```

### Matched-object model

Bilinguals heard leche/perro in Spanish; monolinguals heard milk/dog in
English — same physical objects, collapsed to one label. The four shared
objects are milk, dog, hand, car.

```{r combined-build}
to_object <- function(word) {
  recode(word,
         "leche" = "milk", "perro" = "dog",
         "milk"  = "milk", "dog"   = "dog",
         "hand"  = "hand", "car"   = "car",
         .default = NA_character_)
}
shared_objects <- c("milk", "dog", "hand", "car")

b_matched <- bcorr_clean %>%
  mutate(object = to_object(word), group = "Bilingual") %>%
  filter(object %in% shared_objects) %>%
  select(Participant, group, object, quiet_noise, corrected_acc)

m_matched <- mcorr_clean %>%
  mutate(word = str_remove(Trial.Type, "^(quiet|noise)_"),
         object = to_object(word), group = "Monolingual") %>%
  filter(object %in% shared_objects) %>%
  select(Participant, group, object, quiet_noise, corrected_acc)

matched <- bind_rows(b_matched, m_matched) %>%
  mutate(
    group       = factor(group, levels = c("Monolingual", "Bilingual")),
    quiet_noise = factor(quiet_noise, levels = c("Quiet", "Noise")),
    object      = factor(object),
    Participant = factor(Participant)
  )

cat("Rows:", nrow(matched),
    "| Participants:", nlevels(matched$Participant),
    "| Objects:", nlevels(matched$object), "\n")
matched %>% distinct(Participant, group) %>% count(group)   # N per group
```

```{r combined-model}
m_matched_mod <- lmer(
  corrected_acc ~ group * quiet_noise + (1 | Participant) + (1 | object),
  data = matched
)
summary(m_matched_mod)
```

```{r combined-emmeans}
emm <- emmeans(m_matched_mod, ~ group * quiet_noise)
emm                                   # four estimated cell means (vs 0 = no recognition)
pairs(emm, simple = "group")          # Bilingual vs Monolingual, within each condition
pairs(emm, simple = "quiet_noise")    # Quiet vs Noise, within each group
```

### Matched-object model with SES

This is where group and SES are nearly confounded. Report both models and
watch the `group` coefficient. The collinearity check and VIF quantify how
separable the two are in this subset.

```{r combined-mixed-ses}
matched_ses <- matched %>%
  mutate(Participant = clean_id(Participant)) %>%
  left_join(ses, by = "Participant") %>%
  filter(!is.na(parent_ed_z))

matched_ses %>% distinct(Participant, group, parent_ed) %>%
  group_by(group) %>%
  summarise(n = n(), mean_ed = mean(parent_ed), sd_ed = sd(parent_ed),
            min = min(parent_ed), max = max(parent_ed), .groups = "drop")

c0 <- lmer(corrected_acc ~ group * quiet_noise + (1 | Participant) + (1 | object),
           data = matched_ses)
c1 <- lmer(corrected_acc ~ group * quiet_noise + parent_ed_z +
             (1 | Participant) + (1 | object), data = matched_ses)

summary(c1)
anova(c0, c1)

if (requireNamespace("car", quietly = TRUE)) car::vif(c1)
```

### Within-bilingual SES effect

Does the Quiet–Noise gap shrink or grow with SES? (`quiet_noise:parent_ed_z`
is the term of interest.)

```{r ses-within-bilinguals}
b_noise_ses <- lmer(corrected_acc ~ quiet_noise * parent_ed_z + language + (1 | Participant),
                    data = b_ses_complete)
summary(b_noise_ses)
```

# Habituation Task

79 participants have habituation data (V1 and V2 points combined), all of whom
habituated. We keep one point per participant, preferring V2 when present.

```{r hab-build, include=FALSE}
hab <- read.csv("hab.csv")

hab_filtered <- hab %>%
  mutate(
    base_id = str_remove(Participant, "_V[12]$"),
    visit   = if_else(str_detect(Participant, "_V2$"), 2, 1)
  ) %>%
  group_by(base_id) %>%
  slice_max(visit, n = 1, with_ties = FALSE) %>%
  ungroup()

age  <- read.csv("age.csv")
lena <- read.csv("lena.csv")
mcdi <- read.csv("mcdi.csv")

hab_filtered <- hab_filtered %>%
  left_join(age,  by = "Participant") %>%
  left_join(lena, by = "Participant") %>%
  left_join(mcdi, by = "Participant") %>%
  mutate(language = if_else(str_starts(base_id, "M"), "monolingual", "bilingual")) %>%
  mutate(join_id = clean_id(Participant)) %>%              # SES join on normalized key
  left_join(ses %>% rename(join_id = Participant), by = "join_id") %>%
  select(-join_id)

# age-group bins for plotting
hab_filtered <- hab_filtered %>%
  mutate(
    age_group = case_when(
      age >= 8  & age < 13 ~ "8-12.9",
      age >= 13 & age < 19 ~ "13-18.9",
      age >= 19 & age < 27 ~ "19-26.9",
      TRUE ~ NA_character_
    ),
    age_group = factor(age_group, levels = c("8-12.9", "13-18.9", "19-26.9"))
  )
```

## Missing Data

```{r hab-missing-lena}
# missing LENA
hab_filtered %>% filter(is.na(noise_sum)) %>% pull(Participant)
```

```{r hab-missing-ses}
# missing SES (expect the 5 known-missing: M43EW, B17YT, B23RF, B29NRP, B31ZR)
hab_filtered %>% filter(is.na(parent_ed)) %>% pull(Participant)
```

## Age Distribution

```{r hab-age-hist}
ggplot(hab_filtered, aes(x = age)) +
  geom_histogram(binwidth = 1) +
  geom_rug() +
  theme_classic()
```

## Regressions

```{r hab-split, include=FALSE}
bi   <- hab_filtered %>% filter(language == "bilingual")
mono <- hab_filtered %>% filter(language == "monolingual")
datasets <- list(full = hab_filtered, bilingual = bi, monolingual = mono)
```

### Baseline grid (no SES)

```{r hab-grid}
specs <- crossing(
  outcome = c("nov_score", "hab_trials", "nov1"),
  nesting(
    predictor = c("age * language", "tv", "noise_sum"),
    sample    = c("full", "bilingual", "monolingual")
  )
)

results <- specs %>%
  mutate(fit = pmap(list(outcome, predictor, sample),
                    \(o, p, s) lm(reformulate(p, response = o), data = datasets[[s]]))) %>%
  mutate(tidied = map(fit, tidy)) %>%
  unnest(tidied) %>%
  select(-fit)

results %>% filter(term != "(Intercept)") %>% arrange(p.value)
```

### SES-augmented grid

Every predictor crossed with every sample (dropping language terms in
single-group samples). `n_used` tracks how SES missingness shrinks each model.

```{r hab-grid-ses}
specs_ses <- crossing(
  outcome   = c("nov_score", "hab_trials", "nov1"),
  predictor = c("age * language", "tv", "noise_sum",
                "parent_ed_z",
                "age * language + parent_ed_z",
                "tv + parent_ed_z",
                "noise_sum + parent_ed_z"),
  sample    = c("full", "bilingual", "monolingual")
) %>%
  filter(!(str_detect(predictor, "language") & sample != "full"))

results_ses <- specs_ses %>%
  mutate(fit = pmap(list(outcome, predictor, sample),
                    \(o, p, s) lm(reformulate(p, response = o), data = datasets[[s]]))) %>%
  mutate(n_used = map_int(fit, ~nobs(.x)),
         tidied = map(fit, tidy)) %>%
  unnest(tidied) %>%
  select(-fit)

results_ses %>% filter(term != "(Intercept)") %>% arrange(p.value)
```

### Confirmatory model

Pre-specified after the exploratory grid: full sample, SES entered as a
covariate alongside the predictor of interest (tv). Report N explicitly.

```{r hab-confirmatory}
confirm_df <- hab_filtered %>%
  filter(!is.na(nov_score), !is.na(tv), !is.na(age),
         !is.na(language), !is.na(parent_ed_z))

confirm_mod <- lm(nov_score ~ tv + age + language + parent_ed_z, data = confirm_df)

summary(confirm_mod)
tidy(confirm_mod, conf.int = TRUE)
glance(confirm_mod)
cat("N =", nobs(confirm_mod), "\n")
```

## Novelty Response Plots

```{r hab-nov-box}
ggplot(hab_filtered, aes(x = language, y = nov_score)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  labs(x = "Language Group", y = "Novelty Response Score",
       title = "Novelty Response Scores by Language Group") +
  theme_classic()
```

```{r hab-nov-age}
ggplot(hab_filtered, aes(x = age, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Age (months)", y = "Novelty Response Score", color = "Language Group") +
  theme_classic()
```

```{r hab-nov-agegroup}
ggplot(hab_filtered, aes(x = age_group, y = nov_score, color = language)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.7) +
  labs(x = "Age Group (months)", y = "Novelty Response Score", color = "Language Group") +
  theme_classic()
```

## Habituation + LENA Plots

```{r hab-lena-noise}
ggplot(hab_filtered, aes(x = noise_sum, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "LENA Total Noise", y = "Novelty Response Score", color = "Language Group") +
  theme_classic()
```

```{r hab-lena-tv}
ggplot(hab_filtered, aes(x = tv, y = nov_score, color = language)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "LENA TV Noise", y = "Novelty Response Score", color = "Language Group") +
  theme_classic()
```

<!--
# Parked / not run
The chunks below reference objects (imputed_b, imputed_m) that are not built
in this pipeline. They are kept for reference but set eval=FALSE so they never
break the render. Restore them by supplying the imputed data and flipping eval.

```{r parked-imputed-b, eval=FALSE, include=FALSE}
acc_b <- imputed_b %>%
  group_by(Participant, quiet_noise) %>%
  summarize(Accuracy = mean(Type.Accuracy, na.rm = TRUE), .groups = "drop")

diff_b <- acc_b %>%
  pivot_wider(names_from = quiet_noise, values_from = Accuracy) %>%
  mutate(NoiseDrop = Quiet - Noise) %>%
  arrange(desc(NoiseDrop))
```

```{r parked-imputed-m, eval=FALSE, include=FALSE}
acc_m <- imputed_m %>%
  group_by(Participant, quiet_noise) %>%
  summarize(Accuracy = mean(Type.Accuracy, na.rm = TRUE), .groups = "drop")

diff_m <- acc_m %>%
  pivot_wider(names_from = quiet_noise, values_from = Accuracy) %>%
  mutate(NoiseDrop = Quiet - Noise) %>%
  arrange(desc(NoiseDrop))
```
-->
